---
title: "dplyr_review"
output: html_document
---

## intro to dplyr

dplyr is R package in tidyverse. We can load the package using

```{r remedy001}

#load req package
library(tidyverse)
# or really we can just do that all at once using library(tidyverse) but yea

#preview dataset
head(starwars)



```


### Filtering data
We can filter data using `filter()`. This allows us to subset observations (rows) based on their values (in columns)

Tips:

* Be sure u spell the colnames correctly (and valnames if its a categorical variable). R is CASE SENSITIVE (at least for vars)
* Equality is `==` when comparing observations. (`=` is an assignemnt operator for args)
* You can use < > <= >= to compare numeric or categorical variables (nominal variables are ranked alphabetically, while ordinal variables have a built-in rank)

```{r remedy002}

filter(starwars, hair_color == "blond" | hair_color == "blonde") 

filter(starwars, hair_color == "blond" & species == "Human")

filteredData <- filter(starwars, skin_color == "fair", species == "Human", homeworld == "Tatooine")

# filtering for a specific number or range of numbers

filter(starwars, mass == 75)

filter(starwars, mass >= 75, mass <= 100)

# filtering to choose for everyone except for one value

filter(starwars, mass != 75)

# another way we can specify multiuple arguemnts is using the %in% with a vector

filter(starwars, hair_color %in% c("blond", "blonde", "brown"))

# another reason why %in% is cooler bc it checks all the conditions all at once, when "or" has to go one at a time
# but or is useful when u have two separate conditioins, like hair color OR homeworld

impPlanets <- c('Tatooine', 'Naboo', 'Alderaan')
filter(starwars, homeworld %in% impPlanets)

# we can use > <  <= >= and stoof with strings by doing things alphabetically

#this will filter everyone whos name is before the letter n

filter(starwars, name < "Mace")

# talking more about ranked data
# so often times we have to put in our own order with vectors
# but someitmes it will have it built in, and we can look at that with head(), summary(), class(), unique() etc

head(diamonds) # gives first 6 rows
class(diamonds$cut) # gives u specific type of data kinda like mode
summary(diamonds$cut) # gives you a count of each category or sumamry stats if numeric
summary(diamonds$carat)
head(diamonds$cut) #gives first 6 values and levels

filter(diamonds, cut > "Good")

## Ordering categorical data
unique(starwars$eye_color)

#factor() allows us to force categories into a ordered rank
factor(starwars$eye_color, levels = c(1,2,3,4,5,6,7), labels = c("red", "orange", "gold", "yellow", "green,yellow", "blue", "black", ordered = T))

# find all characters that are shorter than 100 cm

filter(starwars, height < 100, species != "Droid")

# which characters were bron between the years 100 and 200 (inclusive)

filter(starwars, birth_year >= 100, birth_year <= 200)
#faster and cleaner alternative
filter(starwars, between(birth_year, 100, 200))


# which characters weigh over 100kg but are shorter than 185cm?
filter(starwars, mass > 100, height < 185)

### Missing Values, get in the way when we do any summary math calculations
NA > 5 # it doesnt know, NA could be anything
NA + 10 
NA/2
 
NA == NA #this one is weird, these unkown values could be not equal to each other

mass <- NA
10 == mass # returns NA
is.na(mass) # return TRUE

# which characters are missing a hair color?
filter(starwars, is.na(hair_color))

```

### Arranging data

`arrange()` reorders rows. It doesnt remove or add rows. NA values are always at the end when you order by a col

```{r remedy003}
#lowest to highest birth year
arrange(starwars, birth_year)

#highest to lowest birth year
arrange(starwars, desc(birth_year))

#categorical is alphabetical
arrange(starwars, hair_color)

#multiple cols, will go by alphabetical / numerical within each group of the col

arrange(starwars, hair_color, birth_year)

# arrange starwars characters to find the tallest characters and shortest characters

arrange(starwars, height)

# Alphabetize the star wars characters by name

arrange(starwars, name)

# how could u use arrange() to sort all missing values to the start?
arrange(starwars, desc(is.na(height)))
```


### we can select certain columns in the data set

`select()` allows us to retain only certain variables (cols). Doesnt change order, but it removes not named in the args. So filter is for rows, and select is for cols.

```{r remedy004}

select(starwars, hair_color, skin_color, eye_color)

head(starwars)
# we can use a colon to select cols that are in order next to each other
select(starwars, hair_color:eye_color)

# to select everything except a certain col we can use the minus sign
select(starwars, -(hair_color:eye_color))

starwars_no_color <- select(starwars, -(hair_color:eye_color))

# there are many helper functions to select
select(starwars, contains("color"))
select(starwars, ends_with("color")) # there is also starts with
select(starwars, contains("_"))
select(starwars, starts_with("s"), ends_with("color"))

#more helper functions can be accessed with ?select
?select

# in filter, and select, the comma means different things. in select, comma is or, in filter, comma is and

# to rename variables use rename()

rename(starwars, birthYear = birth_year)

# select the homeworld and species for the starwars data. What question might this subset of vars help answer?

select(starwars, homeworld, species)

# come up with as many ways to select these 4 vars
select(starwars, hair_color, eye_color, skin_color, birth_year)
select(starwars, hair_color:birth_year)
select(starwars, ends_with("color"), birth_year)
select(starwars, contains("_"))
select(starwars, -(name:mass), -(sex:starships))
select(starwars, ends_with("color"), ends_with("year"))

```


### adding new columns

`mutate()` adds new columns to the end of ur dataset.

```{r remedy005}
# create smaller dataset to see new vars easier
starwars_small <- select(starwars, height, mass, birth_year)
head(starwars_small)

starwars_small <- mutate(starwars_small, height_m = height/100)
mutate(starwars_small, bmi = mass/(height_m^2))

# we could have also done it straight in one line
mutate(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

# to only keep new columns, use transmute()

transmute(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

# using aggregate functions
# need to use
prop_mass <- mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = TRUE)) 
arrange(prop_mass, desc(proportional_mass))

```
math operations are all vectorized and will apply to each row individually

summary stats go a little differently and apply to the whole col